Multi-scale assessment of a grassland productivity model

### 2020-06-10 manuscript/manuscript.Rmd
some manuscript updates
fix a few references, bold some table values, take URL out of refs except where there is no DOI to link

bibliography: refs
running:
  title: Extended PhenoGrass Evaluation
  author: Taylor & Browning
# This section is mandatory even if you declare that no competing interests are present.
competinginterests: |
  The authors declare no competing interests.
# OPTIONAL:
algorithms: true
# See https://publications.copernicus.org/for_authors/licence_and_copyright.html, normally used for transferring the copyright, if needed.
copyrightstatement: |
  Copyright 2020 The Authors
### The following commands are for the statements about the availability of data sets and/or software code corresponding to the manuscript.
### It is strongly recommended to make use of these sections in case data sets and/or software code have been part of your research the article is based on.
availability:
  codedata: |
    All code and data used in the analysis is available in the repository at https://github.com/sdtaylor/PhenograssReplication, the PhenoGrass model is implemented in a python package https://github.com/sdtaylor/GrasslandModels. Both are archived permanently on Zenodo (link).
  sample: |
    use this section when having geoscientific samples available
authorcontribution: |
  Shawn concieved of the project. Shawn and Dawn wrote the paper.
disclaimer: |
  None
acknowledgements: |
  None
appendix: |
 TODO
output:
  rticles::copernicus_article: default
  bookdown::pdf_book:
    base_format: rticles::copernicus_article # for using bookdown features like \@ref()
---

@@

There are several potential limitations in the underlying productivity models which can drive such a forecast. Process-based models parameterized with observed data have limited transferability beyond the spatial extent from which their training data came [taylor2019]. For any location the most accurate model will be one which was parameterized from locally collected data, yet these site-specific models will not generalize to new locations [@basler2016]. Incorporating more, and diverse, locations into the model building process will allow it to be more generalizable, yet this comes at a cost of decreased proficiency at all locations [@garcia-mozo2008; @basler2016]. Thus there is an optimal extent in the building and subsequent application of productivity models, which depends on a tradeoff between proficiency at the local scale and applicability at the larger scale.

@@

\begin{table}
\caption{Average site-level coefficient of determination (R\textsuperscript{2}) for each vegetation type and ecoregion group parameterized model applied to its respective extent as well as appropriate smaller extents. Bold indicates when a R\textsuperscript{2} was greater than the “acceptable” threshold of 0.65. Values in parentheses represent the average R\textsuperscript{2} in leave-1-out cross validation.}
\begin{tabular}{l|r|r|r|r}
\hline
 & All Site Model & Ecoregion Model & Vegetation Model & Ecoregion+Vegetation Model\\
\hline
All Sites & 0.33 &  &  & \\
\hline
\textbf{E. Temperate Forests} & 0.43 & 0.48 &  & \\
\hline
\hspace{1em}Agriculture & 0.34 & 0.39 & 0.38 & 0.33\\
\hline
\hspace{1em}Grasslands & 0.73 & 0.79 & 0.65 & 0.82\\
\hline
\textbf{Great Plains} & 0.45 & 0.45 &  & \\
\hline
\hspace{1em}Agriculture & 0.22 & 0.15 & 0.17 & 0.18\\
\hline
\hspace{1em}Grasslands & 0.64 & 0.70 & 0.67 & 0.70\\
\hline
\textbf{N. American Deserts} & -0.11 & 0.14 &  & \\
\hline
\hspace{1em}Grasslands & -0.51 & -0.22 & -0.51 & -0.08\\
\hline
\hspace{1em}Shrublands & 0.15 & 0.37 & 0.40 & 0.51\\
\hline
\textbf{N.W. Forests} & 0.25 & 0.58 &  & \\
\hline
\hspace{1em}Grasslands & 0.30 & 0.61 & 0.36 & 0.64\\
\hline
\textbf{All Agriculture} & 0.27 &  & 0.26 & \\
\hline
\textbf{All Grasslands} & 0.43 &  & 0.45 & \\
\hline
\textbf{All Shrublands} & 0.23 &  & 0.39 & \\
\hline
\end{tabular}
\end{table}

@@

The PhenoGrass model performed best in grassland sites embedded within ecoregions. Studies using earlier forms of the model applied it exclusively to grasslands [@choler2010, @choler2011, @hufkens2016], and results here confirm that it performs well in grassland vegetation with two exceptions. The model did not work in the desert grasslands, nor did it generalize well when built using all North American grasslands simultaneously. Grasslands in the N.A. Desert biome coexist with shrubs, resulting in complex water usage dynamics described in more detail below. The pulse-response design of PhenoGrass, which makes it well suited in areas with high cover of perennial grass, is likely not applicable when grasses are interspersed with woody plants.

Shrublands were not well modelled at any extent. Dryland shrubs, representing 8 of the 15 shrubland Phenocams analysed here, coexist with grasses by accessing different pools of soil water [@weltzin2000; @muldavin2008], thus have different responses to precipitation and resulting greenness patterns [@browning2017; @yan2019]. A prior form of the PhenoGrass model was designed to work with dryland shrubs [@ogle2004], yet here PhenoGrass showed little effectiveness. The current model form has only a single soil water pool with fluxes from precipitation and evapotranspiration, thus is not well-suited for representing the deeper water pools which shrubs use. Potential improvements would likely need to incorporate a deep soil water pool, in addition to the shallow, which are each utilized by the respective plant functional groups. This has already been implemented in highly parameterized ecohydrology models [cites] and could potentially be used here to make a more generalized PhenoShrub model to apply across large scales. This approach could also help in modelling N.A. Desert grasslands which coexist among shrubs.

>>>>

bibliography: refs
running:
  title: Extended PhenoGrass Evaluation
  author: Taylor & Browning
# This section is mandatory even if you declare that no competing interests are present.
competinginterests: |
  The authors declare no competing interests.
# OPTIONAL:
algorithms: true
# See https://publications.copernicus.org/for_authors/licence_and_copyright.html, normally used for transferring the copyright, if needed.
copyrightstatement: |
  Copyright 2020 The Authors
### The following commands are for the statements about the availability of data sets and/or software code corresponding to the manuscript.
### It is strongly recommended to make use of these sections in case data sets and/or software code have been part of your research the article is based on.
availability:
  codedata: |
    All code and data used in the analysis is available in the repository at https://github.com/sdtaylor/PhenograssReplication, the PhenoGrass model is implemented in a python package https://github.com/sdtaylor/GrasslandModels. Both are archived permanently on Zenodo (link).
authorcontribution: |
  Shawn concieved of the project. Shawn and Dawn wrote the paper.
acknowledgements: |
  None
appendix: |
  Appendix: TODO
output:
  rticles::copernicus_article: default
  bookdown::pdf_book:
    base_format: rticles::copernicus_article # for using bookdown features like \@ref()
---

@@

There are several potential limitations in the underlying productivity models which can drive such a forecast. Process-based models parameterized with observed data have limited transferability beyond the spatial extent from which their training data came [@taylor2019a]. For any location the most accurate model will be one which was parameterized from locally collected data, yet these site-specific models will not generalize to new locations [@basler2016]. Incorporating more, and diverse, locations into the model building process will allow it to be more generalizable, yet this comes at a cost of decreased proficiency at all locations [@garcia-mozo2008; @basler2016]. Thus there is an optimal extent in the building and subsequent application of productivity models, which depends on a tradeoff between proficiency at the local scale and applicability at the larger scale.

@@

\begin{table}
\caption{Average site-level coefficient of determination (R\textsuperscript{2}) for each vegetation type and ecoregion group parameterized model applied to its respective extent as well as appropriate smaller extents. Bold indicates when a R\textsuperscript{2} was greater than the “acceptable” threshold of 0.65. Values in parentheses represent the average R\textsuperscript{2} in leave-1-out cross validation.}
\begin{tabular}{l|l|l|l|l}
\hline
 & All Site Model & Ecoregion Model & Vegetation Model & Ecoregion+Vegetation Model\\
\hline
All Sites & 0.33 &  &  & \\
\hline
\textbf{E. Temperate Forests} & 0.43 & 0.48 &  & \\
\hline
\hspace{1em}Agriculture & 0.34 & 0.39 & 0.38 & 0.33\\
\hline
\hspace{1em}Grasslands & \textbf{0.73} & \textbf{0.79} & \textbf{0.65} & \textbf{0.82 (0.78)}\\
\hline
\textbf{Great Plains} & 0.45 & 0.45 &  & \\
\hline
\hspace{1em}Agriculture & 0.22 & 0.15 & 0.17 & 0.18\\
\hline
\hspace{1em}Grasslands & 0.64 & \textbf{0.70} & \textbf{0.67} & \textbf{0.70 (0.67)}\\
\hline
\textbf{N. American Deserts} & -0.11 & 0.14 &  & \\
\hline
\hspace{1em}Grasslands & -0.51 & -0.22 & -0.51 & -0.08\\
\hline
\hspace{1em}Shrublands & 0.15 & 0.37 & 0.40 & 0.51\\
\hline
\textbf{N.W. Forests} & 0.25 & 0.58 &  & \\
\hline
\hspace{1em}Grasslands & 0.30 & 0.61 & 0.36 & 0.64\\
\hline
\textbf{All Agriculture} & 0.27 &  & 0.26 & \\
\hline
\textbf{All Grasslands} & 0.43 &  & 0.45 & \\
\hline
\textbf{All Shrublands} & 0.23 &  & 0.39 & \\
\hline
\end{tabular}
\end{table}

@@

The PhenoGrass model performed best in grassland sites embedded within ecoregions. Studies using earlier forms of the model applied it exclusively to grasslands [@choler2010; @choler2011; @hufkens2016], and results here confirm that it performs well in grassland vegetation with two exceptions. The model did not work in the desert grasslands, nor did it generalize well when built using all North American grasslands simultaneously. Grasslands in the N.A. Desert biome coexist with shrubs, resulting in complex water usage dynamics described in more detail below. The pulse-response design of PhenoGrass, which makes it well suited in areas with high cover of perennial grass, is likely not applicable when grasses are interspersed with woody plants.

Shrublands were not well modelled at any extent. Dryland shrubs, representing 8 of the 15 shrubland Phenocams analysed here, coexist with grasses by accessing different pools of soil water [@weltzin2000; @muldavin2008], thus have different responses to precipitation and resulting greenness patterns [@browning2017; @yan2019]. A prior form of the PhenoGrass model was designed to work with dryland shrubs [@ogle2004], yet here PhenoGrass showed little effectiveness. The current model form has only a single soil water pool with fluxes from precipitation and evapotranspiration, thus is not well-suited for representing the deeper water pools which shrubs use. Potential improvements would likely need to incorporate a deep soil water pool, in addition to the shallow, which are each utilized by the respective plant functional groups. This has already been implemented in highly parameterized ecohydrology models [@scanlon2005;@lauenroth2014] and could potentially be used here to make a more generalized PhenoShrub model to apply across large scales. This approach could also help in modelling N.A. Desert grasslands which coexist among shrubs.
